---
title: "lea-FinalReportGraphs"
author: "Lea Collin"
date: "12/9/2018"
output: pdf_document
---

## Main Analysis

```{r, include = FALSE}
library(spotifyr)
# input your own spotify client id and spotify client secret to be able to use Spotify's API
Sys.setenv(SPOTIFY_CLIENT_ID = 'b273df75336c485596266c68d3e7c1c2')
Sys.setenv(SPOTIFY_CLIENT_SECRET = 'e88980163181450eb429c2f3f4723f90')

access_token <- get_spotify_access_token()
```

```{r}
setwd("~/Columbia/EDAV/FinalProject/EDAV-FinalProject")
library(tidyverse)
# this file include track genre and the student's name
lea_class_data <- read_csv("Final_DSI_Playlist.csv", col_names = TRUE)

lea_class_data$rank <- NULL
lea_class_data$album_label <- NULL
lea_class_data$album_name1 <- NULL
lea_class_data$album_id <- NULL
lea_class_data$album_data <- NULL
lea_class_data$album_name2 <- NULL
lea_class_data$album_id1 <- NULL
lea_class_data$album_label1 <- NULL
lea_class_data$album_date <- NULL
lea_class_data$album_date1 <- NULL
```

```{r}
# cleaning up the data a bit

x <- lea_class_data$loudness
lea_class_data$normalized_loudness <- (x-min(x))/(max(x)-min(x))
lea_class_data$normalized_pop <- lea_class_data$track_popularity/100
lea_class_data$normalized_artist_pop <- lea_class_data$artist_popularity/100
x <- lea_class_data$tempo
lea_class_data$normalized_tempo <- (x-min(x))/(max(x)-min(x))
x <- lea_class_data$instrumentalness 
lea_class_data$normalized_instrumentalness <- (x-min(x))/(max(x)-min(x))

display_class_data <- lea_class_data[which(lea_class_data$`Display` == 'Yes'),]

# selecting only the continuous variables from the dataset
cont_var <- lea_class_data %>% select(normalized_pop, danceability, energy, normalized_loudness, speechiness, acousticness, normalized_instrumentalness, valence, normalized_tempo)

names(cont_var) <- c("popularity", "dance", "energy", "loud", "speech", "acoustic", "instrumental", "valence", "tempo")
```

To first begin addressing what distinctive features make a DSI song popular, we decided to first look at how all of these features are generally connected to each other. Most of the features we decided to look at (as they were the most interesting ones), were continuous variables so we made a scatterplot matrix to assess their correlations.

```{r, cache = TRUE}
library(GGally)
scatter_vars <- ggpairs(cont_var, title = "", axisLabels = "show", progress = FALSE)
```

```{r, fig.width = 10, fig.height = 8}
scatter_vars
```

Many of the variables are uncorrelated or very weakly correlated. The strongest positive correlations are between energy and loudness (0.786), valence and energy (0.4), valence and danceability (0.395), and loudness and valence (0.323). It's interesting that track popularity is not really correlated with any of the variables as this makes us wonder what Spotify considers when scoring this. What's interesting is that Spotify claims that it calculates danceability based on the tempo but not only is there almost no correlation between danceability and tempo, but they are even slightly negatively correlated! This might be interesting to look more into as a future direction or given more time. 

We wanted to know more about the most popular songs and artists amoungst EDAV studetns to see if these have any features in common. 

```{r}
top_ten_artists <- as.data.frame(rev(sort(table(lea_class_data$artist_name)))[1:10])
top_ten_artists$Var1 <- as.factor(top_ten_artists$Var1)
top_fifty_songs <- as.data.frame(rev(sort(table(lea_class_data$track_name)))[1:50])
top_fifty_songs$Var1 <- as.factor(top_fifty_songs$Var1)

lea_class_data$`Artist Popularity` <- ifelse(lea_class_data$artist_name %in% top_ten_artists$Var1, "EDAV Top 10", "EDAV Not Top 10")
lea_class_data$`Song Popularity` <- ifelse(lea_class_data$track_name %in% top_fifty_songs$Var1, "EDAV Top 50", "EDAV Not Top 50")

edav_top_50_songs <- lea_class_data[which(lea_class_data$`Song Popularity` == 'EDAV Top 50'), ]
edav_not_top_50_songs <- lea_class_data[which(lea_class_data$`Song Popularity` == 'EDAV Not Top 50'), ]
edav_top_10_artists <- lea_class_data[which(lea_class_data$`Artist Popularity` == 'EDAV Top 10'), ]
edav_not_top_10_artists <- lea_class_data[which(lea_class_data$`Artist Popularity` == 'EDAV Not Top 10'), ]

edav_top_50_songs$playlist_name <- 'EDAV Top 50 Songs'
edav_not_top_50_songs$playlist_name <- 'EDAV Not Top 50 Songs'
edav_top_10_artists$playlist_name <- 'EDAV Top 10 Artists'
edav_not_top_10_artists$playlist_name <- 'EDAV Not Top 10 Artists'
```

To compare EDAV's tastes in 2017 to what's generally popular today and what was popular on Spotify in 2017, we will compare some features to Spotify's 'US Top 50', 'Global Top 50' and 'Top Tracks 2017' playlists.

```{r}
us_top_50 <- get_playlist_audio_features('spotifycharts', '37i9dQZEVXbLRQDuF5jeBp')
global_top_50 <- get_playlist_audio_features('spotifycharts','37i9dQZEVXbMDoHDwVN2tF')
top_2017 <- get_playlist_audio_features('spotify', '37i9dQZF1DX5nwnRMcdReF')

# normalizing all variables to be between 0 and 1
x <- us_top_50$loudness
us_top_50$normalized_loudness <- (x-min(x))/(max(x)-min(x))
us_top_50$normalized_pop <- us_top_50$track_popularity/100
x <- us_top_50$tempo
us_top_50$normalized_tempo <- (x-min(x))/(max(x)-min(x))
x <- us_top_50$instrumentalness 
us_top_50$normalized_instrumentalness <- (x-min(x))/(max(x)-min(x))

x <- global_top_50$loudness
global_top_50$normalized_loudness <- (x-min(x))/(max(x)-min(x))
global_top_50$normalized_pop <- global_top_50$track_popularity/100
x <- global_top_50$tempo
global_top_50$normalized_tempo <- (x-min(x))/(max(x)-min(x))
x <- global_top_50$instrumentalness 
global_top_50$normalized_instrumentalness <- (x-min(x))/(max(x)-min(x))

x <- top_2017$loudness
top_2017$normalized_loudness <- (x-min(x))/(max(x)-min(x))
top_2017$normalized_pop <- top_2017$track_popularity/100
x <- top_2017$tempo
top_2017$normalized_tempo <- (x-min(x))/(max(x)-min(x))
x <- top_2017$instrumentalness 
top_2017$normalized_instrumentalness <- (x-min(x))/(max(x)-min(x))
```

```{r}
excluded_vars <- c('gender', 'user_id', 'location2017', 'nationality', 'undergrad', 'age', 'useFrequency', 'Name', 'Display', 'artist_uri', 'artist_genres', 'Name', 'Song Popularity', 'Artist Popularity', 'artist_popularity', 'artist_num_followers', 'normalized_artist_pop')
edav_top_50_songs <- select(edav_top_50_songs, -excluded_vars)
edav_not_top_50_songs <- select(edav_not_top_50_songs, -excluded_vars)

lea_all_data <- rbind(edav_top_50_songs, us_top_50)
lea_all_data <- rbind(lea_all_data, global_top_50)
lea_all_data <- rbind(lea_all_data, top_2017)
```

To look at the difference in distribution in the continuous variables between all these playlists, we made several density curves and box-plots. Only the most revealing/interesting ones are included within this report. We chose density curves and box-plots as we find these to be the best way to see and compare difference between groups and their continuous variables. 


```{r}
diff_loud_songs <- ggplot(lea_all_data, aes(x = normalized_loudness, fill = playlist_name)) +
  geom_density(alpha = 0.4) + xlab("Loudness")
diff_loud_songs
```

```{r}
diff_tempo_songs <- ggplot(lea_all_data, aes(x = normalized_tempo, fill = playlist_name)) +
  geom_density(alpha = 0.4) + xlab("Tempo")
diff_tempo_songs
```

```{r}
diff_energy_songs <- ggplot(lea_all_data, aes(x = energy, fill = playlist_name)) + 
  geom_density(alpha = 0.4) + xlab("Energy")
diff_energy_songs
```

For energy, we see that there is not such a drastic difference in distribution between the EDAV Top 50 Songs and the other three playlists. However, we clearly see a huge difference in loudness between popular EDAV songs and the popular playlists on Spotify. This is pretty interesting and we also look at this variable as box-plot and with the less popular EDAV songs included below. It also seems that tempo is generally higher for EDAV songs than the others. A box-plot is also included for this below.  

```{r}
lea_all_data <- rbind(lea_all_data, edav_not_top_50_songs)
```

```{r, fig.height = 6}
edav_loud_box <- ggplot(lea_all_data, aes(x=reorder(playlist_name, -1*normalized_loudness, FUN=median), y=normalized_loudness)) + 
  geom_boxplot(fill='#CCCCFF') + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("EDAV, Keep It Down") +
  xlab("Playlist") + ylab("Loudness") + 
  theme_minimal(15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
edav_loud_box
```

From this graph, we see that even when we include the 'less popular' EDAV songs, they are still generally much louder than the songs on the other playlists. Given more time, it could be interesting to look at what songs specifically are causing this trend in the playlists. 

```{r, fig.height = 6}
edav_tempo_box <- ggplot(lea_all_data, aes(x = reorder(playlist_name, -1*normalized_tempo, FUN = median), y = normalized_tempo)) +
  geom_boxplot(fill='#CCCCFF') + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_minimal(15) +
  xlab("Playlist") + ylab("Tempo") + ggtitle("Why are you in such a rush?") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
edav_tempo_box
```

Similar to what we saw from the density curve, we see that EDAV likes slightly faster music than the rest of Spotify users. 
When looking at what makes an artist popular for DSI, we first wanted to see if the popularity of the artist on Spotify matched the popularity of the artist in DSI. To look at this, we have a density curve of artist popularity (Spotify's metric) between popular and less popular artists in DSI.

```{r}
lea_all_data <- rbind(edav_top_10_artists, edav_not_top_10_artists)

diff_art_pop <- ggplot(lea_all_data, aes(x = normalized_artist_pop, fill = playlist_name)) + 
  geom_density(alpha = 0.4) + xlab("Artist Popularity")
diff_art_pop
```

From this graph, we see that the artists that are popular with DSI students are also considered by Spotify to be popular, so DSI's aritst tastes match closely with the rest of Spotify users. 

```{r}
diff_art_dance <- ggplot(lea_all_data, aes(x = normalized_tempo, fill = playlist_name)) + 
  geom_density(alpha = 0.4) + xlab("Tempo")
diff_art_dance
```

From the graph above, we see something that is slightly inconsistent with what we saw with the most popular DSI songs. We see here that the most popular DSI artists actually have songs that are generally slower paced than less popular songs in DSI. This is a fairly interesting discrepancy. 

#### Second Part of Main Analysis

To explore both how similar DSI students are in their music preferences and the key ways in which they differ, we thought it would first be interesting to see which students like songs and artists that have been deemed 'popular' by Spotify. 


```{r, fig.height = 8, fig.width = 10}
avg_class_data <- display_class_data %>% group_by(Name) %>% summarise_at(vars(normalized_artist_pop, normalized_pop, danceability, energy, valence, normalized_loudness, speechiness, acousticness, normalized_instrumentalness, normalized_tempo), funs(mean(., na.rm = TRUE)))

names(avg_class_data) <- c("Name", "Artist Popularity", "Track Popularity", "Danceability", "Energy", "Valence", "Loudness"," Speechiness", "Acousticness", "Instrumentalness", "Tempo")

tidy_avg <- avg_class_data %>% select(Name, `Track Popularity`, `Artist Popularity`) %>% 
  arrange(`Track Popularity`) %>% 
  mutate(Name = factor(Name, levels = .$Name)) %>% tidyr::gather(key = 'Category', value = 'Percentage', -Name)

pop_cd <- ggplot(tidy_avg, aes(x = Percentage, y = Name, color = Category)) +
          geom_point() + ylab("") + theme_minimal() +
          ggtitle("Hammaad and Lea Are Basic AF") +
        scale_color_manual(values=c("red", "blue")) +
    theme(legend.title=element_text(size=15), legend.text=element_text(size=13))
pop_cd
```

