---
title: "Final Project"
author: "Danielle Su (ds3731)"
date: "12/1/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE, 
                      cache = TRUE)
```

```{r, include=FALSE}
library(spotifyr)
library(tidyverse)
Sys.setenv(SPOTIFY_CLIENT_ID = 'XXXXXXX')
Sys.setenv(SPOTIFY_CLIENT_SECRET = 'XXXXXXXX')
access_token <- get_spotify_access_token()
```

```{r}
class_df <- read_csv('Final_DSI_Playlist.csv')
```

In examining the differences between students in our DSI cohort, we wanted to see who listens to the most obscure music, or the most "indie", and who listens to the most popular must, or the most "mainstream". The best variable to do this with is Spotify's track popularity variable, this is a number assigned from 0 to 100 that is calculated based on an algorithm that takes the total number of streams for that tracks and weights more recent streams more heavily. The exact formula for this number is unknown, but it is our best measure in the dataset for how popular a song is. One caveat to this variable is that we pulled the track popularity when we built our dataset, and these playlists are from 2017, so the track_popularity is indicative of how popular that song is today rather than in 2017 when the student was listening to it. Therefore the results could also be interpreted as, who listens to the music that stays popular throughout a year's worth of time.

Below we created 2 sets of boxplots - they are boxplots of the distribution of their playlist's popularity variables for each student in our dataset. You can see based on the first set which has track popularity that Lea listens to the most popular music in terms of having the highest median track popularity value and Eric has the lowest median track popularity. Additionally, from the boxplots you can see that some students such as Mert and Anonymous 1 have very large ranges whereas James and Hammaad's ranges are rather small in comparison.

Additionally I made the same plot for the artist popularity variable, ordered in the same way as track popularity, so it is easy to compare. From Spotify's description, the artist popularity variable is determined by the track popularity so it should be a similar graph. However, when you compare, you can see that all ranges are smaller and in general they are on the higher end of the popularity variable. This could be explained by a few reasons, maybe artist popularity distributions in general are higher, or students could be listening to low popularity tracks but from artists that do have very high popularity tracks, or it could be a factor of the time difference. Perhaps the tracks that students listened to in 2017 are no longer very popular, but the artist has new tracks in 2018 that have brought the popularity variable up.

```{r fig.height=5}
mytheme <- theme_minimal(16) +
    theme(plot.title = element_text(hjust = 0.5))

# Boxplots of track popularity for each student, ordere by highest median track popularity to lowest
ggplot(class_df, aes(x=fct_reorder(Name, track_popularity), y=class_df$track_popularity)) +
  geom_boxplot(fill="#CCCCFF") +
  scale_x_discrete("") +
  labs(y="Track Popularity", x="DSI-er", title="Who is the most indie and the most mainstream") +
  coord_flip() +
  mytheme


# Boxplots ordered the same as track_popularity (for comparison) but showing artist popularity
ggplot(class_df, aes(fct_reorder(Name, track_popularity), artist_popularity)) +
  geom_boxplot(fill="#CCCCFF") +
  labs(y="Artist Popularity", x="DSI-er", title="Does this look the same when showing artist popularity?") +
  coord_flip() +
  mytheme

```

The next few graphs try to use the survey inputs to identify any trends within those survey answers. While the number of tracks in our dataset is sizeable (2400), they only come from 24 students, and when separating that into categories based on survey answers, it is hard to really identify any trends. We were mainly interested in the gender variable and the age variable. For gender, the distribution in the dataset is 8 females and 16 males. For age, the distribution in the dataset is age 21: 1, age 22: 4, age 23: 4, age 24: 5, age 25: 5, age 26: 2, age 28: 2. Therefore while the graphs using age are still interesting, it is hard to say any general trends, since if you're talking about 21 year olds it is really just that 1 specific person. Additionally the ages are all in the 21-28 range, so I wouldn't expect a huge amount of difference in their listening habits, but you can see some distinctions.

Below we made a mosaic plot of age and track popularity, which is binned into 4 quartiles since the original number is likely a percentile. You can see from the mosaic plot that there isn't much of a relationship between age and listening to popular or not popular tracks. It looks like 22 year olds listen to the most popular tracks, but this could also be because there is only 1 21 year old and at least 4 22 year olds. Therefore nothing conculsive can really be said about age and popularity. If given more time, we could collect a wider range of samples and test this again.

```{r}
library(forcats)
library(ggmosaic)

popular_df <- class_df %>% select(track_popularity, age)

# Binning the track popularity into quartiles
df <- popular_df %>% select(track_popularity, age) 
df$track_popularity[df$track_popularity < 25] <- "[1-25)"
df$track_popularity[df$track_popularity >= 25 & df$track_popularity < 50] <- "[25-50)"
df$track_popularity[df$track_popularity >= 50 & df$track_popularity < 75] <- "[50-75)"
df$track_popularity[df$track_popularity >= 75] <- "[75-100)"

df$age <- factor(df$age)
df$track_popularity <- factor(df$track_popularity)

ggplot(df) +
  geom_mosaic(aes(x = product(age), fill=track_popularity), na.rm=FALSE)

```
  
Next, we wanted to explore the album data we were able to pull and seeing if there are any trends based on the year the album was release and therefore the song was release. 

First, we explored album dates in relation to gender of the student. Below is a bar graphs of the number of frequency of the track and the decade the song was released, faceted by gender. You can see that the distribution looks similar but there is a little cluster of songs released in the 70s for the males - classic rock songs, no doubt. Since there are a lot more males than females in our dataset, I graphed the same thing again in terms of percentages. In that next graph you can now see that the females in our group have over 80% of their songs coming out of the 2010s decade. So it seems like everyone likes more recent music, but especially DSI females!

```{r}

# Getting total Male/Female counts to calculate percentages
gender_count <- class_df %>% group_by(gender) %>% summarize(total=n())
female_count <- as.integer(gender_count[1,2])
male_count <- as.integer(gender_count[2,2])

# Creating new dataframe with grouped values by decade and gender
class_df_grouped <- class_df %>% group_by(album_decade, gender) %>% summarize(Freq=n()) %>% mutate(Percent= ifelse(gender=='Female', 100*(Freq/female_count), ifelse(gender=='Male', 100*(Freq/male_count), 0)))

# Bar chart with counts
ggplot(class_df_grouped, aes(x=album_decade, y=Freq)) +
  geom_histogram(fill="thistle", stat="identity") +
  facet_wrap(~gender)

# Bar char with percentages - you can really see the percentage of females listening to recent music is high
ggplot(class_df_grouped, aes(x=album_decade, y=Percent)) +
  geom_histogram(fill="thistle", stat="identity") +
  facet_wrap(~gender) +
  labs(xlab='Decade of Album Release', title='Everyone likes more recent music, but especially DSI Females!')


```

Since the majority of the songs came out of alubms in the 2000s I wanted to do a deeper dive into the 2000s to see if there are any trends within years of the 2000s.

In terms of gender, the distributions of the 2000s look similarly skewed. However, males have fatter tails in terms of having a higher percentage from 2000 to 2012, so while they are listening to more recent music, maybe they like to listen to throwback songs a little more than females. Additionally, now that we are zooming into the 2000s you can see that of the 2010s the majority of the songs are from 2016-2017, so either songs that came out the year of the playlist or songs that came out the year right before it. 

I would also like to note that there are some album release dates in 2018, even though this was a playlist of top songs played in 2017. This is due to the fact that the way Spotify has tracked their album release dates, it was easiest to take the release date from the actual album, whereas it might have shown up on the playlist because the single came out in 2017, but the album did not come out until 2018.

```{r}
# Deeper Dive into the 2000s
year2000 <- class_df %>% filter(album_year >= 2000)

# Creating new dataframe with grouped values by year and gender
year2000_grouped <- year2000 %>% group_by(album_year, gender) %>% summarize(Freq=n()) %>% mutate(Percent= ifelse(gender=='Female', 100*(Freq/female_count), ifelse(gender=='Male', 100*(Freq/male_count), 0)))

# Male/Female distributions are pretty similar
ggplot(year2000_grouped, aes(x=album_year, y=Percent)) +
  geom_histogram(fill="thistle", stat="identity") +
  facet_wrap(~gender) +
  labs(xlab='Year of Album Release', title='Similar distributions, but males like to throwback a little more')
```

Since the gender distributions were similar, even though age cannot be generalized, we wanted to see if age had any trends with album release year. The below graphs are faceted by age and you can see that generally as you increas in age, the skew to 2016-2017 is less and less extreme and students listen to more and more songs released in the early 2000s. In order to better compare, we made a plot with density curves of each age. Here you can see similar trends, but you can really see that 26 year olds have the lowest density around 2016-2017 and are much more evenly spread out than the others. You can also see that 21 and 22 year olds have the lowest percentages of songs from 2000 to about 2007.

```{r}
# Creating new dataframe with grouped values by year and gender
age21 <- length(which(class_df$age == '21'))
age22 <- length(which(class_df$age == '22'))
age23 <- length(which(class_df$age == '23'))
age24 <- length(which(class_df$age == '24'))
age25 <- length(which(class_df$age == '25'))
age26 <- length(which(class_df$age == '26'))
age27 <- length(which(class_df$age == '27'))
age28 <- length(which(class_df$age == '28'))

year2000_grouped <- year2000 %>% group_by(album_year, age) %>% summarize(Freq=n()) %>% 
  mutate(Percent= ifelse(age=='21', 100*(Freq/age21), ifelse(age=='22', 100*(Freq/age22), ifelse(age=='23', 100*(Freq/age23), ifelse(age=='24', 100*(Freq/age24), ifelse(age=='25', 100*(Freq/age25), ifelse(age=='26', 100*(Freq/age26),ifelse(age=='27', 100*(Freq/age27),ifelse(age=='28', 100*(Freq/age28), 0)))))))))

# Bar char with percentages
ggplot(na.omit(year2000_grouped), aes(x=album_year, y=Percent)) +
  geom_histogram(fill="thistle", stat="identity") +
  facet_wrap(~age) +
  labs(xlab='Year of Album Release', title='Year')

# Density curves
year2000clean <- na.omit(subset(year2000, select=c(album_year, age)))
Student_Age <- as.factor(year2000clean$age)
ggplot(year2000clean, aes(x = album_year, color=Student_Age)) +
  geom_density() +
  ggtitle("Age/Year of album distribution") +
  labs(x = "Year of Album Release", y = "Density") +
  theme(plot.title = element_text(hjust = 0.5))
```


We were curious to see how all these factors: age, track popularity, and album release date come together, so we plotted them on a scatterplot with date on the x-axis and popularity on the y-axis. The points are colored by the age of the student. In accordance with the previous plots, most of the songs are clustered around 2016-2017, and they range the full spectrum of track popularity with a big cluster in the 30-75 range and a small clusterin the 0-15 range. Generally, it looks like if younger students listen to older songs they at least have a track popularity of 40 or above. The only students that listen to tracks released before 1991 and are below 40 in popularity are 24 and 25 year olds, you can see this even better in the interactive and subsetted graph described below.

In order to see specifics of these outliers, and to better visualize the graph, we subset the data so it only includes the top 25 songs in each person's playlist, presumably these are their top 25 favorite songs of 2017. Then, we used plotly and displayed the track name and artist name so we could see what exactly these points represented. The oldest song in this subset is "I Fall in Love too Easily" by Chet Baker with a popularity level of 62. The highest track popularity for a song that came out before 2000 is 81 and that is for "Under the Bridge" by Red Hot Chili Peppers. The highest track popularity overall is 89 and is the song Jocelyn Flores by XXXTENTACION. The lowest track popularity for a song that came out before 2000 is 0 and is the song "Come on Home" by Joan Armatrading.

```{r}
splot_theme <- theme_classic(12) + theme(plot.title = element_text(hjust = .5))
library(viridis)
library(plotly)

# Scatterplot with all songs comparing album release date and track popularity, colored by age
class_df_clean <- na.omit(subset(class_df, select=c(album_date, age, track_popularity)))
Age <- as.factor(class_df_clean$age)
ggplot(class_df_clean, aes(album_date, track_popularity)) + 
    geom_point(aes(color= Age), alpha = .7, size = 2, stroke=0) +
    scale_color_viridis_d() +
    splot_theme + 
    labs(title= 'How does album release date relate to track popularity?', subtitle = "All Tracks",
         x = 'Album Release Date', y = 'Track Popularity Levels')

# Same as above, but with only the top 25 songs per student's playlist
top50_df <- na.omit(subset(class_df, rank <=25, select=c(album_date, age, track_popularity, track_name, artist_name)))
Age <- as.factor(top50_df$age)
g<- ggplot(top50_df, aes(album_date, track_popularity, text=paste("Song:",track_name, "Artist:",artist_name))) + 
    geom_point(aes(color = Age), alpha = .7, size = 2, stroke=0) +
    scale_color_viridis_d() +
    splot_theme + 
    labs(title= 'How does album release date relate to track popularity?', subtitle = "Top 50 Tracks of Each Playlist",
         x = 'Album Release Date', y = 'Track Popularity Levels')

# Made interactive so we can see what those outlier songs are that were released a long time ago but have high track popularity
# Hover over points to see what song it is and the artist in addition to date of release and track pouplarity
ggplotly(g)

```

